Determinism, Complexity, and Predictability in Computer Performance

### 2014-05-30 response.tex

\documentclass[12pt]{article}

\usepackage{amsmath}

\title{Response to the Reviewers}
\author{Joshua Garland, Ryan James, and Elizabeth Bradley}
\date{\today}

\begin{document}

\maketitle

\section*{Response to the First Referee}

\emph{In this manuscript the authors present a way to explore the relationship
between predictability and complexity across a broad (but not complete) array of
forecast strategies such as random walk (last value), na\"ive (average value),
regression based (ARIMA) and a nonlinear method based on state space
reconstruction (LMA). They use weighted permutation entropy as a criterion/index
of predictive structure in the time series and illustrate their approach on
eight real-world datasets derived from three different systems (computer
performance traces).}

\emph{Overall the manuscript is well written and clearly understandable. The
analysis seems to be carried out carefully and the conclusions drawn are quite
reasonable. While I don't get the impression that this is a breakthrough paper I
can still recommend the manuscript for publication in PRE if the issues listed
below are addressed thoroughly.}

\noindent\emph{Major issues:}

\emph{I don't get the reason for normalizing the three methods by the random
walk method (for example in Table I). As the authors write themselves this makes
the indicator vulnerable to a bias caused by the high influence of this
(arbitrarily selected) reference method. Why not just use a non-referential
estimator, which then would also allow to compare the inherent predictability of
different systems (as it is now the values obtained for each systems are
strongly influenced by the random walk estimate on that particular system and
thus can not be compared). This will also change the results of the fitting (the
authors explicitly exclude one value because of the effect caused by the
normalization). And it is not that the random walk predictor is the one standard
method that all other methods have to compete against.}

\emph{In many parts the article reads almost like a review. There is a lot of
unnecessary redundancy (the entropy rate is quite low). In particular, the main
point (``The aim of the paper is not ... but ...'') is repeated far too many
times. The paper while easy to read could be shortened considerably without any
loss of information.}

\noindent\emph{Minor issues:}

\emph{Page 1: Why mention the `halting problem'? I don't see a straightforward
connection to the problem at hand. The only connection is `undecidability' but
there are many other examples for that.}

\emph{Page 10: `error between the predictions and the true continuations' -->
`error of the predictions with respect to the true continuations' Figure 6 on
Page 11: It would be more intuitive to turn this into a `table of figures'
with the four methods as row labels on the left and the three systems as column
labels at the top. Also: Caption `for forecast of ... and all four prediction
strategies'.}

\emph{First equation on Page 14: Shouldn't the index $j$ run from $i+1$ to
$i+\ell$ (instead of from $i$ to $i+\ell$)? You normalize by $\ell$ (not
$\ell+1$) values.}

\emph{Page 19: ``There are some exceptions: [...] Conversely, forecast errors
for the \texttt{col\_major} and \texttt{dgesdd$_5$} signals are higher than the
corresponding WPEs suggest --- except for the nonlinear LMA predictor.'' This is
not really an exception. As mentioned by the authors before, in general, the
result only holds under the assumption that a reasonably good model has been
used for the prediction. But for ill-matched datasets and models like in the
cases mentioned above this is just not the case so it is not a surprise that the
forecast errors are higher.}

\section*{Response to the Second Referee}

\emph{The problem of time series prediction is considered in the paper of J.
Garland et al. Namely, the authors suggest to quantify predictability in advance
(before using any concrete prediction scheme) based on the complexity estimate
(weighted permutation entropy, WPE). They hypothesize that WPE is related to the
best possible prediction error in a simple way. For several benchmark time
series (reflecting variations of a computer performance during execution of
different tasks), they obtain estimates of WPE and prediction error (the least
one over four techniques), and build a linear regression using seven or eight
resulting data points. They suppose that the obtained linear function can be
used to ``predict'' the best prediction error from the model-free WPE estimate.
Thus, the authors finally assume (as could be seen from the last section) that
WPE and optimal prediction error for all systems are related via the same
universal linear function with fixed coefficients. This is what I can understand
from the entire paper, though the problem is not so definitely described in the
Introduction.}

\emph{I think that the possibility to estimate the best prediction error in
advance from a relatively simple WPE characteristic would be interesting and
useful, if it could hold true for a wide class of systems. Such a question would
deserve a careful study as that started in the presented work. However, I have a
number of critical remarks and strongly doubt that the conjecture of the authors
can be valid for a reasonably wide class of systems (not even speaking of ``all
systems''). Other aspects of the paper are not so important, in my view, despite
some of them are fresh and interesting (e.g. taking a computer as an object of
prediction). My critical remarks are listed below.}

\emph{1) Introduction is too long, but the main problem is not formulated
definitely enough. I see that the main point is the approximately linear (or at
least one-to-one) relationship between WPE and the best prediction error. This
aim is not claimed explicitly in the Introduction. All other aspects of the
manuscript do not seem important enough to be published in a separate article.}

\emph{2) In particular, two primary findings mentioned on page 2 (Introduction)
are formulated quite vaguely. The first part of the first finding reads ``(i)
complexity of a noisy real-value time series is quantifiable by permutation
entropy''. However, it seems obvious from the previous research, including that
cited by the authors. The authors do not suggest PE or WPE here, they exploit
them as previously known approaches. It was previously known that PE is an
estimate of KS entropy as the authors also state. It was well known that KS
entropy relates to predictability and, in this sense, it is a measure of
complexity (see e.g. the work [G. Boffetta, M. Cencini, M. Falcioni, and A.
Vulpiani, ``Predictability: a way to characterize complexity'' // Physics
Reports 356 (2002) 367–474] which is not cited by the authors). Thus, the first
part of the first finding does not provide any new information. The second part
of the first finding reads ``(ii) complexity of the noisy real-valued time
series is correlated with prediction accuracy of an appropriate predictor''.
However, it seems also obvious and directly follows from the definition of KS-
entropy and its discussion in many papers including [Boffetta et al.] cited
above.}

\emph{3) By the way, the authors call ``fully complex'' the white noise process,
i.e. they equate predictability and complexity. However, they do not even
mention another approach to the notion of complexity which ascribes low
complexity to the white noise [the works e.g. by C.R. Shalizi and J. Crutchfield
carried out in the Santa Fe Institute from where one of the authors is]. It
would be appropriate to mention that ambiguity of the notion of complexity if
the latter plays an important role in many formulations of the authors.}

\emph{4) The second finding reads ``The way information is generated and
processed internally by a system plays a crucial role in the success of
different forecasting schema - and in the choice of which one is appropriate for
a given time series''. This statement also looks trivial. Indeed, an optimal
predictor for a linear stochastic system and low-dimensional nonlinear
deterministic system are quite different (an AR model versus a local or global
nonlinear model) due to different properties of the original systems. Thus,
there is nothing to be proven here. Probably, the authors imply their particular
benchmark signals (reflecting a computer behavior) but they do not claim that.
Thus, both primary findings seem to provide nothing new.}

\emph{5) The benchmark system - a computer as an object of prediction - seems to
be an unexpected and, probably, interesting choice but just as an additional
illustration of a statement which should be first shown for well-controlled and
well-understood paradigmatic systems (like low-dimensional maps or ordinary
differential equations). Indeed, it is not known in advance what error one
should expect for such a complicated object using a certain predictive
technique, what technique is optimal or close to optimal, etc. If the main point
of the research is the relationship between WPE and the best prediction error
(if any), selection of such a complicated object as computer performance
variations only introduces its own difficulties into the problem.}

\emph{6) The description of the choice of the benchmark system in the
Introduction is too long. On the other hand, if the computer signals are the
main interest of the work, then it should be claimed explicitly. Then, the paper
should be rewritten and submitted to a kind of engineering journals where it
might well be of interest.}

\emph{7) As for the choice of the four prediction techniques, some of them seem
superfluous. E.g. naive approach is just a global AR model of the zero order.
Random-walk is similar to the AR model of the first order where the coefficient
is not estimated by the least-squares but set equal to 1 (or this is an ARIMA
model with the first difference and ``zero ARMA part''). Inefficiencies of the
ARIMA model as compared to those choices in the paper can represent an
inappropriate procedure for the order selection rather than indicate
inappropriateness of the entire ARIMA approach. Thus, linear ARIMA and nonlinear
LMA might be sufficient (in combination with mathematical benchmark examples) to
represent close-to-optimal prediction errors.}

\emph{8) Conditions for inefficiency of simple prediction schemes (like naive
and random-walk) discussed in the paper look rather obvious as well and do not
give new information. Is it the purpose of the paper to discuss when and where a
concrete simple prediction scheme fails?}

\emph{9) A better justification of the prediction error metrics could be also in
order if this work is continued. At least, comparison of prediction errors
quantified with different metrics seems necessary. Especially, taking into
account the following major concerns.}

\noindent\emph{General remarks.}

\emph{I) Strict relationship between WPE and prediction error applicable to all
systems is hardly possible. In particular, there is a notion of epsilon-entropy
which tends to the KS entropy if epsilon (the cell size) tends to zero (see e.g.
[Boffetta et al.] cited above). For stochastic systems, the KS entropy is
infinite and epsilon-entropy varies strongly with the scale epsilon. Seemingly,
the permutation entropy estimate in this case will strongly depend on the choice
of the word length L. The authors concentrate on L = 6 and obtain coefficients
of the regression between WPE and prediction error only for that choice and for
their particular signals. However, other systems and other choices of L may
easily lead to other regression coefficients so that deciding about the best
possible prediction error from WPE seems impossible in general. It is good if
such a decision is possible for a certain narrow class of systems. However, the
authors do not even discuss to which class of systems their results apply.}

\emph{II) PE is invariant under an invertible nonlinear change of variables. The
prediction error measured as the authors suggest is not invariant. Thus, such a
simple change of variables can lead, at least, to a change in regression
coefficient between WPE and prediction error. Thus, the authors should also
discuss to which variable for that class of systems their result apply. I think
that the fact that the authors observe a certain regression line at all seems to
be due to small number of different signals they consider (only 7 or 8 different
data points, such a small number of points may lie approximately near a straight
line by chance).}

\emph{To summarize, the paper shows that the authors are well familiar with the
entire topic and highly qualified, they overview some well-known facts (WPE,
various prediction techniques, etc) and ``invent'' an unexpected test system
(computer performance data), they report about their attempts to play around
with all that staff. However, their results do not seem to provide any new
information to a specialist. Probably, this work can be continued to get some
more reliable results about concrete relationship between WPE and prediction
error for certain classes of systems (however, I doubt that it is possible to
derive any simple and widely applicable relationship like the linear function
suggested by the authors). Seemingly, it should then be a new paper which can be
hardly considered as a revised version of the present manuscript.}

\end{document}

### 2014-05-30 model.tex
beating this thing into PRE shape...
put a few responses in. things in red have not actually been done yet

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
Some example plots of data from
Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d.png}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

    \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gcc3dipc.png}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure} 
  \\
  \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svd53dipc2.png}
    \caption{\svdfive}
    \label{fig:svdfiveEmbedding}
  \end{subfigure}%   
%

     %\includegraphics[width=\textwidth]{figs/colipc3d.png}
     \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ts} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
 \label{fig:embedding}
 \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.  
%

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error between the predictions and the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

@@

  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colRWForecast.png}
    \caption{\col\\ random walk }
    \label{fig:colRW}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccRWForecast.png}
    \caption{\gcc\\ random walk }
    \label{fig:gccRW}
  \end{subfigure}% 
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast.png}
    \caption{\svdfive\\ random walk}
    \label{fig:svd5RW}
  \end{subfigure}%      
  \\  
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colMeanForecast.png}
    \caption{\col\\ na\"ive }
    \label{fig:colMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccMeanForecast.png}
    \caption{\gcc\\ na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}% 
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast.png}
    \caption{\svdfive\\ na\"ive }
    \label{fig:svd5MEAN}
  \end{subfigure}%      
  \\
    \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colARIMAForecast.png}
    \caption{\col \\ ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast.png}
    \caption{\gcc \\ ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast.png}
    \caption{\svdfive \\ ARIMA }
    \label{fig:svd5ARIMA}
  \end{subfigure}%    
  \\

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  
% On this type of plot, a perfect prediction lies exactly on the
% diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

@@

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
  \begin{center}
  \begin{tabular}{|c|c|c|c|c|}
  \hline Signal & LMA  & ARIMA & na\"{i}ve  & Weighted Permutation \\ 
  &  MASE & MASE  & MASE &  Entropy \\ \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

@@

%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
%shown in Figure~\ref{fig:lmacol}. 
%
%\begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
%    \caption{\col }
%    \label{fig:lmacol}
%  \end{subfigure}%  
%  
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
%    \caption{\gcc}
%    \label{fig:lmagcc}
%  \end{subfigure} 
%  \begin{subfigure}{\columnwidth}
%%    \includegraphics[width=\columnwidth]{figs/%colPredShortTS}
%    \caption{\svdfive}
%    \label{fig:lmasvdfive}
%  \end{subfigure}% 
%  \caption{{\color{red} [[If Liz thinks we should %include these I actually need to generate this %figure]]}An LMA-based forecast of
%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
%       predicted values, respectively; vertical bars %show where these
%       values differ.}\label{fig:lmapredictions}
% \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{figs/colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{figs/colPredShortTS}
%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
%On the other hand, if we use the same approach to %forecast the
%processor efficiency (IPC) of the \gcc time series, %the
%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.

>>>>

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
Some example plots of data from
Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
 \begin{figure}
   \centering
\begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colipc3d}
    \caption{\col }
    \label{fig:colEmbedding}
  \end{subfigure}%  \\

    \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
    \caption{\gcc}
    \label{fig:gccEmbedding}
  \end{subfigure}
  \\
  \begin{subfigure}{\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
    \caption{\svdfive}
    \label{fig:svdfiveEmbedding}
  \end{subfigure}%
%

     %\includegraphics[width=\textwidth]{colipc3d}
     \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ts} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
 \label{fig:embedding}
 \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.
%

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 360
traces described in Section~\ref{sec:methods}, then calculate the
error of the predictions with respect to the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

@@

  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colRWForecast}
    \caption{\col\\ random walk }
    \label{fig:colRW}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
    \caption{\gcc\\ random walk }
    \label{fig:gccRW}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
    \caption{\svdfive\\ random walk}
    \label{fig:svd5RW}
  \end{subfigure}%
  \\
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
    \caption{\col\\ na\"ive }
    \label{fig:colMEAN}
  \end{subfigure}%
   \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
    \caption{\gcc\\ na\"ive }
    \label{fig:gccMEAN}
  \end{subfigure}%
     \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
    \caption{\svdfive\\ na\"ive }
    \label{fig:svd5MEAN}
  \end{subfigure}%
  \\
    \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
    \caption{\col \\ ARIMA}
    \label{fig:colARIMA}
  \end{subfigure}
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
    \caption{\gcc \\ ARIMA }
    \label{fig:gccARIMA}
  \end{subfigure}%
  \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
    \caption{\svdfive \\ ARIMA }
    \label{fig:svd5ARIMA}
  \end{subfigure}%
  \\

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
% On this type of plot, a perfect prediction lies exactly on the
% diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

@@

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
\caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
  \hline\hline Signal & LMA MASE  & ARIMA MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
  \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

@@

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline\hline
  \end{tabular*}
  \end{center}
 \label{default}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7.  This is in accordance with the visual appearance of
the corresponding images in Figure~\ref{fig:forecast-example}.  While
the correspondence between MASE score and the visual appearance of
these kinds of plots is clear cut in this case, that is not always so.
The plots of LMA predictions of \col and \svdfive both lie near the
diagonal, for instance, but the corresponding MASE scores were very
different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20 and
1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a double-edged
sword.  Here, it facilitates comparisons across signals of different
types and lengths, but its particular pathologies must be taken into
account, as discussed at more length in Section~\ref{sec:results}.

@@

%---works quite well on the trace in Figure~%\ref{fig:ipc}, as%
%shown in Figure~\ref{fig:lmacol}.
%
%\begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\col }
%    \label{fig:lmacol}
%  \end{subfigure}%
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\gcc}
%    \label{fig:lmagcc}
%  \end{subfigure}
%  \begin{subfigure}{\columnwidth}
%%    \includegraphics[width=\columnwidth]{%colPredShortTS}
%    \caption{\svdfive}
%    \label{fig:lmasvdfive}
%  \end{subfigure}%
%  \caption{{\color{red} [[If Liz thinks we should %include these I actually need to generate this %figure]]}An LMA-based forecast of
%       processor-efficiency performance trace from the%\col ,\gcc and \svdfive generating processes.  Red %circles and blue $\times$s are the true and
%       predicted values, respectively; vertical bars %show where these
%       values differ.}\label{fig:lmapredictions}
% \end{figure}
%
%\begin{figure}[htbp]
%  \centering
%   \includegraphics[width=\textwidth]{colPredShortTS}
%    \caption{A forecast of the last 4,000 points of the %signal in
%      Figure~\ref{fig:ipc} using an LMA-based strategy on %the
%      embedding in Figure~\ref{fig:embedding}.  Red circles %and blue
%      $\times$s are the true and predicted values, %respectively;
%      vertical bars show where these values differ.}
%\label{fig:lmacol}
%\end{figure}
%
%\begin{figure}[htbp]
%  \centering
%    \includegraphics[width=\textwidth]{colPredShortTS}
%     \caption{{\color{red} [[actually need to generate this %figure]]}An LMA-based forecast of the last 4,000 points of a
%       processor-load performance trace from the \gcc
%       benchmark.  Red circles and blue $\times$s are the %true and
%       predicted values, respectively; vertical bars show %where these
%       values differ.}
%\label{fig:lmagcc}
%\end{figure}
%
%On the other hand, if we use the same approach to %forecast the
%processor efficiency (IPC) of the \gcc time series, %the
%prediction is far less accurate; see Figure~%\ref{fig:lmagcc}.
%Figure~\ref{fig:lmasvdfive} is clearly a better %prediction than Figure~\ref{fig:gccLMA} but not nearly as good as the \col forecast.
